Collective SU($n$) spin model dynamics

### 2020-08-12 sun_phases/sun_dynamics_notes.tex
work out mean single-particle observables in the non-interacting limit

We first consider a general quadratic spin Hamiltonian of the form
\begin{align}
  H = \f1N \sum_{\substack{\mu,\nu,\rho,\sigma\\i<j}}
  h^{\mu\nu i}_{\rho\sigma j} s_{\mu\nu i} s_{\rho\sigma j}
  + \sum_{\mu,\nu,i} \epsilon_{\mu\nu i} s_{\mu\nu i},
  \label{eq:spin}
\end{align}
where $\mu,\nu\in\set{-I,-I+1,\cdots,I}$ with $I\equiv\p{n-1}/2$ index
orthogonal states of an $n$-level spin; $i,j\in\ZZ_N$ index one of $N$
spins; $h^{\mu\nu i}_{\rho\sigma j}$ and $\epsilon_{\mu\nu i}$ are
scalars; and $s_{\mu\nu i}\coloneqq\op{\mu}{\nu}_i$ is a transition
operator for spin $i$.  If interactions are SU($n$) symmetric then
$h^{\mu\nu i}_{\rho\sigma k} = h_{ik}
\delta_{\mu\sigma}\delta_{\nu\rho}$, in which case
\begin{align}
  H = \f1N \sum_{\substack{\mu,\nu,\rho,\sigma\\i<j}}
  h_{ij} \Pi_{ij}
  + \sum_{\mu,\nu,i} \epsilon_{\mu\nu i} s_{\mu\nu i},
  &&
  \Pi_{ij} \equiv \sum_{\mu,\nu} s_{\mu\nu i} s_{\rho\sigma j}.
  \label{eq:spin_sun}
\end{align}

@@

We can essentially prepare two types of initial product states
$\ket\Psi\equiv\ket\psi^{\otimes N}$.  First and foremost, we can
prepare arbitrary polarized states defined by the polar and azimuthal
angles $\theta,\phi$:
\begin{align}
  \ket{\theta,\phi}
  \equiv e^{i\phi s_\z} e^{i\theta s_\y} \ket\up
  = \sum_\mu \zeta_\mu\p{\theta} e^{-\i\mu\phi} \ket\mu,
  \label{eq:polarized_state}
\end{align}
where $\ket\up\equiv\ket{I}$ is a state polarized along the $z$ axis,
and\footnote{The expansion of $\ket{\theta,\phi}$ in
  \eqref{eq:polarized_state} and \eqref{eq:polarized_proj} is acquired
  by representing an $n$-level spin by $n-1$ qubits restricted to the
  (Dicke) manifold of permutationally symmetric states.}
\begin{align}
  \zeta_\mu\p{\theta} \equiv { 2I \choose I+\mu }^{1/2}
  \cos\p{\f{\theta}{2}}^{I+\mu} \sin\p{\f{\theta}{2}}^{I-\mu}.
  \label{eq:polarized_proj}
\end{align}
For even $n$, we can also prepare the following superposition of
polarized states:
\begin{align}
  \ket{\DS}
  \equiv \f1{\sqrt{2}}
  \p{e^{\i\gamma} \ket{\alpha,\beta}
    + e^{-\i\gamma} \ket{\alpha,-\beta}}
  = \sum_\mu \zeta_\mu\p{\alpha}
  \times \sqrt{2}\, \cos\p{\mu\beta-\gamma} \ket\mu
  \label{eq:double_state}
\end{align}
with
\begin{align}
  \alpha = \f{\pi}{2} + \arcsin\p{\f13},
  &&
  \beta = \f{\pi}{3},
  &&
  \gamma = -\f{2\pi}{3}\, I.
\end{align}

@@

We consider the single-particle Hamiltonian
\begin{align}
  H = -\f{J}{2} \sum_{\mu,q} \cos\p{q + \mu\xi} s_{\mu\mu q},
\end{align}
where $q$ is a quasi-momentum (in units with lattice spacing $a=1$);
$\mu\in\set{-I,-I+1,\cdots,I}$ is a nuclear spin projection onto a
quantization axis, with $I\equiv\p{n-1}/2$ the total nuclear spin;
$\xi$ is a spin-orbit-coupling angle, and $J$ is the single-particle
bandwidth (proportional to the tunneling rate, $J/4$).  As a dynamical
order parameter, we consider the operator
\begin{align}
  \O\p{t}
  \equiv e^{\i t H} \p{\f{\v S^\dag \c\v S}{N^2}} e^{-\i t H}
  = \f1{N^2} \sum_{\mu,\nu,p,q} s_{\mu\nu p} s_{\nu\mu q} \times
  e^{\i tJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}},
\end{align}
where
\begin{align}
  \eta_{\mu\nu p}
  \equiv -\f12 \sp{\cos\p{p+\mu\xi} - \cos\p{p+\nu\xi}}
  = \sin\p{\xi_{\mu\nu}^-} \sin\p{p+\xi_{\mu\nu}^+},
  &&
  \xi_{\mu\nu}^\pm \equiv \f{\mu\pm\nu}{2}\,\xi.
\end{align}
Specifically, we are interested in the expectation value
\begin{align}
  \bk{\Psi|\O\p{t}|\Psi}
  &= \f1{N^2} \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \sum_{p\ne q} e^{\i tJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}} + \f{n}{N} \\
  &= \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \abs{\chi_{\mu\nu}\p{\xi}}^2 + \f{n-1}{N},
\end{align}
where
\begin{align}
  \chi_{\mu\nu}\p{\xi} \equiv \f1N \sum_q e^{\i tJ\eta_{\mu\nu q}}
  = \f1N \sum_q e^{\i tJ\sin\p{\xi_{\mu\nu}^-} \sin\p{q+\xi_{\mu\nu}^+}}.
\end{align}
As $N\to\infty$, this sum is well approximated by an integral:
\begin{align}
  \chi_{\mu\nu}\p{\xi} \approx \f1{2\pi} \int_{-\pi}^\pi \dd\theta\,
  e^{\i tJ\sin\p{\xi_{\mu\nu}^-} \sin\p{\theta+\xi_{\mu\nu}^+}}
  = \J_0\sp{tJ\sin\p{\xi_{\mu\nu}^-}},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.  To
leading order in $N\gg1$, we thus find that
\begin{align}
  \bk{\Psi|\O\p{t}|\Psi}
  \approx \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \J_0\sp{tJ\sin\p{\xi_{\mu\nu}^-}}^2.
  \label{eq:O_time}
\end{align}

@@

The constant (time independent) terms in \eqref{eq:O_time} have
$\mu=\nu$ (in which case all $\xi_{\mu\nu}^-=0$), so up to $O(1/N)$
corrections
\begin{align}
  \bk{\Psi|\O_{\t{const}}|\Psi}
  \approx r_\psi \equiv \sum_\mu \abs{\bk{\mu|\psi}}^4,
  \label{eq:O_const}
\end{align}
which is sometimes called the inverse participation ratio of
$\ket\psi$ with respect to the basis $\set{\ket\mu}$ of states with
definite projection onto the $z$ axis.  Figure \ref{fig:limiting_vals}
shows several values of $r_\psi$ for different $n$.

At ``short'' times $t$ for which $tJ\sin\p{I\xi}\ll1$, we can expand
the Bessel function in \eqref{eq:O_time} as
$\J_0\p{\tau}^2\approx1-\tau^2/2+O\p{\tau^4}$ to get
\begin{align}
  \bk{\Psi|\O\p{t}|\Psi}
  \approx 1 - \f12 \p{tJ}^2 \sum_{\mu,\nu}
  \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2 \sin\p{\xi_{\mu\nu}^-}^2.
\end{align}
In the limit of a small angle $\xi$ for which $I\xi\ll1$, we can
further simplify
\begin{align}
  \bk{\Psi|\O\p{t}|\Psi}
  \approx 1 - \f14 \var_\psi\p{Z} \p{JI\xi}^2 \times t^2,
  \label{eq:small_time_angle}
\end{align}
where
\begin{align}
  \var_\psi\p{Z} \equiv \bk{\psi|Z^2|\psi} - \bk{\psi|Z|\psi}^2,
  &&
  Z \equiv \sum_\mu \f{\mu}{I} \op{\mu}.
\end{align}
Note that the result in \eqref{eq:small_time_angle} is valid for all
$t\ll\p{JI\xi}^{-1}$, which may be a long time given the small-angle
assumption $I\xi\ll1$.  Figure \ref{fig:limiting_vals} shows several
values of $\var_\psi\p{Z}$ for different $n$.

@@

We now consider the opposite, strongly-interacting limit of a 1D
model,

>>>>

We first consider a general quadratic spin Hamiltonian of the form
\begin{align}
  H = \f1N \sum_{\substack{\mu,\nu,\rho,\sigma\\i<j}}
  h^{\mu\nu i}_{\rho\sigma j} s_{\mu\nu i} s_{\rho\sigma j}
  + \sum_{\mu,\nu,i} \epsilon_{\mu\nu i} s_{\mu\nu i},
  \label{eq:spin}
\end{align}
where $\mu,\nu\in\set{I,I-1,\cdots,-I}$ with $I\equiv\p{n-1}/2$ index orthogonal states of an $n$-level spin; $i,j\in\ZZ_N$ index one of $N$ spins; $h^{\mu\nu i}_{\rho\sigma j}$ and $\epsilon_{\mu\nu i}$ are scalars; and $s_{\mu\nu i}\coloneqq\op{\mu}{\nu}_i$ is a transition operator for spin $i$.
If interactions are SU($n$) symmetric then $h^{\mu\nu i}_{\rho\sigma k} = h_{ik} \delta_{\mu\sigma}\delta_{\nu\rho}$, in which case
\begin{align}
  H = \f1N \sum_{\substack{\mu,\nu,\rho,\sigma\\i<j}}
  h_{ij} \Pi_{ij}
  + \sum_{\mu,\nu,i} \epsilon_{\mu\nu i} s_{\mu\nu i},
  &&
  \Pi_{ij} \equiv \sum_{\mu,\nu} s_{\mu\nu i} s_{\rho\sigma j}.
  \label{eq:spin_sun}
\end{align}

@@

We can essentially prepare $N$-fold tensor products $\ket\Psi\equiv\ket\psi^{\otimes N}$ of three types of initial product states $\ket\psi$.
First and foremost, we can prepare states $\ket{\mu}$ of definite spin projection $\mu=\set{I,I-1,\cdots,-I}$ onto a quantization axis.
In addition, we can prepare arbitrary polarized states defined by the polar and azimuthal angles $\theta,\phi$:
\begin{align}
  \ket{\theta,\phi}
  \equiv e^{i\phi s_\z} e^{i\theta s_\y} \ket\up
  = \sum_\mu \zeta_\mu\p{\theta} e^{-\i\mu\phi} \ket\mu,
  \label{eq:polarized_state}
\end{align}
where $\ket\up\equiv\ket{I}$ is a state polarized along the $z$ axis, and\footnote{The expansion of $\ket{\theta,\phi}$ in \eqref{eq:polarized_state} and \eqref{eq:polarized_proj} is acquired by representing an $n$-level spin by $n-1$ qubits restricted to the (Dicke) manifold of permutationally symmetric states.}
\begin{align}
  \zeta_\mu\p{\theta} \equiv { 2I \choose I+\mu }^{1/2}
  \cos\p{\f{\theta}{2}}^{I+\mu} \sin\p{\f{\theta}{2}}^{I-\mu}.
  \label{eq:polarized_proj}
\end{align}
Finally, for even $n$ we can also prepare the following superposition of polarized states:
\begin{align}
  \ket{\DS}
  \equiv \f1{\sqrt{2}}
  \p{e^{\i\gamma} \ket{\alpha,\beta}
    + e^{-\i\gamma} \ket{\alpha,-\beta}}
  = \sum_\mu \zeta_\mu\p{\alpha}
  \times \sqrt{2}\, \cos\p{\mu\beta-\gamma} \ket\mu
  \label{eq:double_state}
\end{align}
with
\begin{align}
  \alpha = \f{\pi}{2} + \arcsin\p{\f13},
  &&
  \beta = \f{\pi}{3},
  &&
  \gamma = -\f{2\pi}{3}\, I.
\end{align}

@@

We consider the single-particle Hamiltonian
\begin{align}
  H = -\f{J}{2} \sum_{\mu,q} \cos\p{q + \mu\xi} s_{\mu\mu q},
\end{align}
where $q$ is a quasi-momentum (in units with lattice spacing $a=1$); $\mu\in\set{I,I-1,\cdots,-I}$ is a nuclear spin projection onto a quantization axis, with $I\equiv\p{n-1}/2$ the total nuclear spin; $\xi$ is a spin-orbit-coupling angle, and $J$ is the single-particle bandwidth (proportional to the tunneling rate, $J/4$).
As dynamical order parameters, we consider the operators
\begin{align}
  s_{\mu\nu} \equiv \f1N \sum_q s_{\mu\nu q},
  &&
  s^2 \equiv \v s^\dag\cdot \v s
  = \f1{N^2} \sum_{\mu,\nu,p,q} s_{\mu\nu p} s_{\nu\mu q},
\end{align}
where $\v s_q\equiv\sum_{\mu,\nu} s_{\mu\nu q}\uv e_{\mu\nu}$ is a vector of normalized generators of $U(n)$ for spin $q$, and $\uv e_{\mu\nu}$ is a unit vector of $\RR^{n^2}$.
In the Heisenberg picture, these operators at time $t$ are
\begin{align}
  s_{\mu\nu}\p{t} = \f1N \sum_q s_{\mu\nu q}
  \times e^{\i t J \eta_{\mu\nu q}},
  &&
  s^2\p{t} = \f1{N^2} \sum_{\mu,\nu,p,q} s_{\mu\nu p} s_{\nu\mu q}
  \times e^{\i tJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}},
\end{align}
where
\begin{align}
  \eta_{\mu\nu q}
  \equiv -\f12 \sp{\cos\p{q+\mu\xi} - \cos\p{q+\nu\xi}}
  = \sin\p{\xi_{\mu\nu}^-} \sin\p{p+\xi_{\mu\nu}^+},
  &&
  \xi_{\mu\nu}^\pm \equiv \f{\mu\pm\nu}{2}\,\xi.
\end{align}
In particular, we are interested in the expectation values
\begin{align}
  \bk{\Psi|s_{\mu\nu}\p{t}|\Psi}
  = \f1N \bk{\psi\op{\mu}{\nu}\psi} \sum_q e^{\i t J \eta_{\mu\nu q}}
  = \bk{\psi\op{\mu}{\nu}\psi} \chi_{\mu\nu}\p{\xi},
\end{align}
and
\begin{align}
  \bk{\Psi|s^2\p{t}|\Psi}
  &= \f1{N^2} \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \sum_{p\ne q} e^{\i tJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}} + \f{n}{N} \\
  &= \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \abs{\chi_{\mu\nu}\p{\xi}}^2 + \f{n-1}{N},
\end{align}
where
\begin{align}
  \chi_{\mu\nu}\p{\xi} \equiv \f1N \sum_q e^{\i tJ\eta_{\mu\nu q}}
  = \f1N \sum_q e^{\i tJ\sin\p{\xi_{\mu\nu}^-} \sin\p{q+\xi_{\mu\nu}^+}}.
\end{align}
As $N\to\infty$, this sum is well approximated by an integral:
\begin{align}
  \chi_{\mu\nu}\p{\xi} \approx \f1{2\pi} \int_{-\pi}^\pi \dd\theta\,
  e^{\i tJ\sin\p{\xi_{\mu\nu}^-} \sin\p{\theta+\xi_{\mu\nu}^+}}
  = \J_0\sp{tJ\sin\p{\xi_{\mu\nu}^-}},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
To leading order in $N\gg1$, we thus find that
\begin{align}
  \bk{\Psi|s_{\mu\nu}\p{t}|\Psi}
  &\approx \bk{\psi\op{\mu}{\nu}\psi}
  \J_0\sp{tJ\sin\p{\xi_{\mu\nu}^-}},
  \\
  \bk{\Psi|s^2\p{t}|\Psi}
  &\approx \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \J_0\sp{tJ\sin\p{\xi_{\mu\nu}^-}}^2.
  \label{eq:ops_time}
\end{align}

@@

The constant (time independent) terms in \eqref{eq:ops_time} have $\mu=\nu$ (in which case all $\xi_{\mu\nu}^-=0$), so up to $O(1/N)$ corrections
\begin{align}
  \bk{\Psi|s_{\mu\nu}^{\t{const}}|\Psi}
  \approx \abs{\bk{\mu|\psi}}^2,
  &&
  \bk{\Psi|s^2_{\t{const}}|\Psi}
  \approx r_\psi \equiv \sum_\mu \abs{\bk{\mu|\psi}}^4,
  \label{eq:ops_const}
\end{align}
where the quantity $r_\psi$ is sometimes called the inverse participation ratio of $\ket\psi$ with respect to the basis $\set{\ket\mu}$ of states with definite projection onto the $z$ axis.
Figure \ref{fig:limiting_vals} shows several values of $r_\psi$ for different $n$.

At ``short'' times $t$ for which $tJ\sin\p{I\xi}\ll1$, we can Taylor expand the Bessel functions in \eqref{eq:ops_time} about $t=0$ to get
\begin{align}
  \bk{\Psi|s_{\mu\nu}\p{t}|\Psi}
  &\approx \bk{\psi\op{\mu}{\nu}\psi}
  \times \sp{1 - \f14 \p{t J}^2 \sin\p{\xi_{\mu\nu}^-}^2}, \\
  \bk{\Psi|s^2\p{t}|\Psi}
  &\approx 1 - \f12 \p{tJ}^2 \sum_{\mu,\nu}
  \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2 \sin\p{\xi_{\mu\nu}^-}^2.
\end{align}
In the limit of a small angle $\xi$ for which $I\xi\ll1$, we can further simplify
\begin{align}
  \bk{\Psi|s_{\mu\nu}\p{t}|\Psi}
  &\approx \bk{\psi\op{\mu}{\nu}\psi}
  \times \sp{1 - \f1{16} \p{\f{\mu-\nu}{I}}^2
    \p{JI\xi}^2 \times t^2}, \\
  \bk{\Psi|s^2\p{t}|\Psi}
  &\approx 1 - \f14 \var_\psi\p{Z} \p{JI\xi}^2 \times t^2,
  \label{eq:small_time_angle}
\end{align}
where
\begin{align}
  \var_\psi\p{Z} \equiv \bk{\psi|Z^2|\psi} - \bk{\psi|Z|\psi}^2,
  &&
  Z \equiv \sum_\mu \f{\mu}{I} \op{\mu}.
\end{align}
Note that the result in \eqref{eq:small_time_angle} is valid for all $t\ll\p{JI\xi}^{-1}$, which may be a long time given the small-angle assumption $I\xi\ll1$.
Figure \ref{fig:limiting_vals} shows several values of $\var_\psi\p{Z}$ for different $n$.

@@

We now consider the strongly-interacting limit of a 1D model, in which the Hamiltonian is
\begin{align}
  H = -\f{U}{N} \v S^\dag\cdot \v S
  - \f{J}{2} \sum_{\mu,q} \cos\p{q + \mu\xi} s_{\mu\mu q}
\end{align}
with $J\ll U$.
